How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-07-24 doc/article.tex
waiting for seth comments

There are a number of ways in which past extinction patterns might be used to model present risk. The simplest case assumes that relationships between predictors (hereafter covariates) and extinction risk have been constant through time. A more complex but more realistic case allows relationships between covariates and extinction risk to vary through time, consistent with evidence for temporal variation in extinction selectivity regime. Finally, an important consideration is the degree to which species’ geographic range trajectories exhibit deterministic versus Markovian behavior \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}. In the former case, knowledge of the specific past trajectory of a species – whether its range has expanded or contracted from some point in the past to the present – might help to improve assessments of its current risk. In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past).

@@

We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). One of the few multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig. \ref{fig:oos_ins_diff}). There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, but none are as long as that for Radiolaria.

@@

We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future.
%One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig. \ref{fig:auc_compare}). However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). The model with the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time. However, the out-of-sample forecasts from this model are biased, indicating that it is overfit to the data \ref{fig:fold_auc}). The model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the model that allows covariate effects to vary over time but does not include historical covariates.

We note that both including historical covariates such as change in geographic range and allowing covariate effects to vary over time are different ways of encoding information from the past. Including historical covariates in a model but not allowing covariate effects to vary over time encodes the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, on the other hand, does not explicitly encode ``the past'' into our model but instead models how covariate effects vary over time which allows the past to be implicitly included in our model. By modeling this variation, forecasts made for future extinction events can allow for a wide range of potential effects of our covariates on predicting extinction. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.

@@

To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts, suggesting that including historical information via either explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data.

>>>>

There are a number of ways in which past extinction patterns might be used to model present risk. The simplest case assumes that relationships between predictors (hereafter covariates) and extinction risk have been constant through time. A more complex but more realistic case allows relationships between covariates and extinction risk to vary through time, consistent with evidence for temporal variation in extinction selectivity regime. Finally, an important consideration is the degree to which species’ geographic range trajectories exhibit deterministic versus Markovian behavior \citep{Liow2007,Kiessling2013,Foote2007b,Pigot2012}. In the former case, knowledge of the specific past trajectory of a species – whether its range has expanded or contracted from some point in the past to the present – might help to improve assessments of its current risk. In the latter case only the current geographic range of the species would convey useful information about current and future risk (although that assessment would still be based on the relative extinction risk of species that had similar ranges at different points in the geological past). In all cases, we use information about the past to predict the future, the question is wheter and how much historical information (e.g. the histories of species over time) improves our ability to forecast future extinction events.

@@

We compared the difference in AUC estimates from the out-of-sample forecasts to the AUC estimates from in-sample forecasts by subtracting the in-sample AUC estimates from the out-of-sample AUC estimates (Fig. \ref{fig:oos_ins_diff}); a difference in AUC close to 0 indicates complete congruence between the in-sample and out-of-sample forecasts. A positive difference indicates that out-of-sample forecasts actually outperform in-sample forecasts, whereas negative difference indicates poorer out-of-sample performance than in-sample forecast. Divergences between out-of-sample and in-sample forecasts are rare and tend not to cluster in time, consistent with the broad visual congruence between the in-sample and out-of-sample performance (Fig. \ref{fig:auc_taxon_time}, \ref{fig:fold_auc_taxon_time}). An example multimillion year pattern indicating significantly poorer out-of-sample forecast performance than in-sample forecast performance is for Radiolaria based on the VP model between 35 Mya and approximately 28 Mya (Fig. \ref{fig:oos_ins_diff}). There exist similar periods of worse out-of-sample forecasting performance for other combinations of taxonomic group and predictive model, for example the CP model has worse out-of-sample forecasting than in-sample for the last 5 million years of the Cenozoic. In general, however, most out-of-sample and in-sample forecasts are almost identical.

@@

We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). These results confirm that past extinction patterns can provide valuable information about which extant species are most threatened with extinction in the near geological future, and that some historical information does not degrade our ability to forecast future extinction risk. To reiterate, all of our models are fit to past extinction events from the Cenozoic and covariates like geographic range or global climate state are associated with these extinctions; our models are conditioned on the past. Some of our models, however, encode historical information such as how covariate effects have varied over time or the change in a species geographic range over time (Table \ref{tab:model_def}). 
%One of the most striking results is that the models that include either the historical covariates (CP model) or allows covariate effects to vary over time (V model) produce less biased out-of-sample forecasts than the model which includes both historical covariates and allows covariate effects to vary over time (VP model) (Fig. \ref{fig:auc_compare}). However, we might expect a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

Three of the four models we evaluated are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). Our VP model had the best in-sample forecasting results includes the historical covariates and allows all covariate effects to vary over time. However, the out-of-sample forecasts from this model are biased, indicating that it is overfit to the data \ref{fig:fold_auc}). The CP model that includes historical covariates such as geographic range trajectory yields out-of-sample forecasts with nearly identical results to the V model that allows covariate effects to vary over time but does not include historical covariates.

While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple ways to encode historical information which describes the species over time such as change in geographic range or previous climate state or by allowing covariate effects to vary over time. Models that include the historical covariates (e.g. change in geographic range) but do not allow covariate effects to vary over time (i.e. CP model) encode the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects vary over time, thus implicitly encoding historical information about the species in our models. By modeling the variation in covariate effects over time, forecasts made for future extinction events are conditioned on a wider range of potential covariate effects which can improve our models flexibility when forecasting extinction in the novel environmental conditions we might expect in the future. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.

@@

To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data, and that historical information. Additionally, including historical information via explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.
